---
title: "Data"
author: "R LAS"
date: "December 14, 2015"
output: ioslides_presentation
---

# Matrices

## Defining a matrix

Defining a matrix is straightforward. Use the option byrow=TRUE to fill across the rows.
```{r}
A<-matrix(1:6,nrow=2)
A
```
```{r}
A<-matrix(1:6,nrow=2,byrow=TRUE)
A
```

## Recycling

Recycling will occur if needed and you'll get a warning.

```{r}
A<-matrix(1:6,nrow=2,ncol=5)
A

```

## Matrices are their own class

```{r}
class(A)
attributes(A)
attr(A,"dim")
```

***

```{r}
nrow(A)
dim(A)[1]
ncol(A)
dim(A)[2]

```

## Selecting individual elements, rows and columns.

```{r}
A[2,2]
A[1,]
A[,2]
```


***

Note that the class of a single row or column reverts to a vector. To keep it as a matrix, use `drop=FALSE`.

```{r}
class(A[1,])
class(A[1,,drop=FALSE])
A[1,,drop=FALSE]
```

## Submatrices

```{r}
set.seed(6345789)
x<-sample(1:20,size=9,replace=TRUE)
m1<-matrix(x,nrow=3)
m1
```

***
Select the 1st and 3rd rows and the 2nd and 3rd columns.
```{r}
m1[c(1,3),c(2,3)]
```

```{r}
m1[-2,-1]
```

## Using Row and Column Names

```{r}
rownames(m1)
colnames(m1)
rownames(m1)<-c("row1","row2","row3")
colnames(m1)<-c("col1","col2","col3")
m1[c("row1","row2"),c("col2","col3")]
```



## Character matrices and mixed matrices

```{r}

B<-matrix(c("a","b","c","d"),nrow=2);B
C<-list(1,2,3,"a")
dim(C)<-c(2,2);C
```


## Using rbind and cbind to add rows and columns

```{r}
A<-matrix(1:6,nrow=2)
B<-rbind(A,7:9)
B
dim(B)
```

***

```{r}
C<-cbind(A,10:11)
C
dim(C)
```



## Matrix Operations

```{r}
A<-matrix(1:6,nrow=2)
3*A
A+A
```

## Matrix Operations

`A^2` squares each element. The transpose of A is given by `t(A)`.
```{r}
A<-matrix(1:6,nrow=2)
A^2
t(A)
```

## Matrix Multiplication

Remember that when multiplying two matrices AB the number of columns of A has to equal the number of rows in B.

```{r}
B<-matrix(c(-1,6,8,7,4,1),nrow=3)
C<-A %*% B
dim(C)
C
# B %*% A will give an error
```

## Solving Equations

Solve the following system of equations.
$$ x + y =8$$
$$ 3x + 2y=18$$
```{r}
(A<-matrix(c(1,1,3,2),nrow=2,byrow=TRUE))
(B<-matrix(c(8,18),nrow=2))
```

## Solution

First, check that the determinant is non-zero so that it has a solution.
```{r}
det(A)
```
Then use `solve`.
```{r}
solve(A,B)
```

## Computing the inverse of a matrix

`solve` also gives the inverse of a matrix.
```{r}
solve(A)
solve(A)%*%B
```

## Eigenvalues

`eigen` computes eigenvalues and eigenvectors. 
```{r}
eigen(A)
```

## Diagonal Matrices

`diag` extract the diagonal of a matrix and also creates a diagonal matrix, depending on its argument.
```{r}
diag(A)
diag(c(2,4,6,8))
```

## Row/Column Operations

`rowsums, colsums, rowmeans, colmeans` give the sums and means of rows and columns.

```{r}
(A<-matrix(1:6,nrow=2))
rowSums(A)
colSums(A)
```

***
```{r}
rowMeans(A)
colMeans(A)
```



## Exercise: Create a matrix that gives the following table. 

Team      | Wins   | Losses  | Pct
----------|--------|---------|----
Toronto   |    93  |   69    |0.574
New York  |    87  |   75    |0.537
Baltimore |    81  |   81    |0.500
Tampa Bay |    80  |   82    |0.494
Boston    |    78  |   84    |0.481


## Solution

```{r}
teams<-c("Toronto","New York","Baltimore","Tampa Bay","Boston")
wins<-c(93,87,81,80,78)
losses<-c(69,75,81,82,84)
```

```{r}
alEast<-matrix(c(wins,losses),nrow=5,byrow=FALSE)
alEast
```

## Add row and column names
```{r}
rownames(alEast)<-teams
colnames(alEast)<-c("Wins","Losses")
alEast
```

## Add winning percentage

```{r}
Pct<-round(alEast[,"Wins"]/(alEast[,"Wins"]+alEast[,"Losses"]),3)
alEast<-cbind(alEast,Pct)
alEast
```



## What if we want the team names to be a column in the matrix?
```{r}
alEast<-matrix(c(teams,wins,losses),nrow=5,byrow=FALSE)
alEast
```

## Try to coerce column to numeric
```{r}
class(alEast[,2])
as.numeric(alEast[,2])
alEast[,2]<-as.numeric(alEast[,2])
alEast
```

## Use `cbind` to put them together

```{r}
alEast<-cbind(teams,wins,losses)
alEast
class(alEast)
```

We need a matrix-like structure that allows columns to be of different classes.

# Data Frames

##
Data frames allow for columns with different classes.
```{r}
alEast<-data.frame(Team=teams,W=wins,L=losses)
alEast
class(alEast)

```

## Classes are different.

```{r}
sapply(alEast,class)
```

Use `stringsAsFactors=FALSE` to use characters instead of factors.

```{r}
alEast<-data.frame(Team=teams,W=wins,L=losses,
                   stringsAsFactors=FALSE)
sapply(alEast,class)
```

## Selecting rows and columns

We can access rows and columns the same way as matrices. 

```{r}
alEast[,"Team",drop=FALSE]
class(alEast[,"Team",drop=FALSE])
```

## We can also access them like elements of a list.

```{r}
alEast$W

alEast[[2]]

dim(alEast)
```

## Attributes of a data frame
```{r}
attributes(alEast)
```


## Lahman Baseball Database

R package created from [Sean Lahman's](http://seanlahman.com/) baseball database. We'll look at the Teams dataframe.
```{r}
library(Lahman)
class(Teams)
dim(Teams)
```

## Column and row names

Instead of `colnames`, we just use `names`.

```{r}
names(Teams)
```

***

Instead of `rownames`, we use `row.names`. In this case they're just the row numbers. We also have commands to get the number of rows and columns.

```{r}
head(row.names(Teams))
nrow(Teams)
ncol(Teams)
```

## Exploring data frames

You've already seen `head` and `tail`. They let you see the top and bottom of the data frame. An additional argument lets you give the number of rows to be shown.

```{r}
head(Teams,6)
```

***
```{r}
tail(Teams,4)
```

## Summary

```{r}
summary(Teams)
```

## Structure
```{r}
str(Teams)
```

## Data Frame Operations

* select - pick columns by name
* filter - keep rows matching criteria
* mutate - add new variables
* arrange - reorder rows
* summarize - reduce variables to values

## Select

Let's get a data frame similar to theone we just created byselecting the Year, League, Team, Division, Wins, and Losses.

```{r}
myTeams<-Teams[,c("yearID","lgID","teamID","divID","W","L")]
head(myTeams)
```

***
Use `str` to see what the league ID and division ID's are.
```{r}
str(myTeams)
unique(myTeams$divID)
```

## Filter

Find the 2014 AL East standings.
```{r}
alEast2014<-myTeams[myTeams$yearID==2014 & myTeams$lgID=="AL" 
                    & myTeams$divID=="E",]
alEast2014
```

***
Let's eliminate the columns that are the same.
```{r}
alEast2014<-alEast2014[,c("teamID","W","L")]
names(alEast2014)<-c("Team","Wins","Losses")
alEast2014
```


## Mutate

Add the percentage of wins.

```{r}
alEast2014$Pct<-round(alEast2014$Wins/(alEast2014$Wins+alEast2014$Losses),3)
alEast2014
```

## Arrange

List the teams from highest to lowest winning percentage. (Note the - indicates decreasing.)

```{r}
alEast2014[order(-alEast2014$Pct),]
```


## Summarize

Determine the total wins for each of 6 divisions in 2014.

```{r}
mlb2014<-subset(Teams,yearID==2014)
mlb2014<-mlb2014[,c("lgID","divID","W")]
d1<-aggregate(W~lgID+divID,data=mlb2014,sum)
```


# Getting Data in and out of R

## Basics

There's a manual, "R Data Import/Export", located at C:/Program Files/R/R-X.Y.Z/doc/manual/R-data.pdf

### Getting Data In
`read.csv` and `read.fwf` are straightforward to use. Pay attention to the option `stringsAsFactors`.
`read.table` gives you more options. `readLines` will read in a text file and returns a character vector of length the number of lines read.

### Getting Data Out

`write.csv` and `write.table` export data in formats that are read by, e.g. Excel and Access.

## readr and data.table Packages

### readr

The `readr` package has functions for reading and writing csv files (`read_csv` and `write_csv`.) These functions are faster than those included in base R. 

You can read about it at [http://blog.rstudio.org/2015/10/28/readr-0-2-0/](http://blog.rstudio.org/2015/10/28/readr-0-2-0/)

### data.table

`data.table` has a function `fread` that is a "Fast and friendly file finagler." By default it returns a data.table but will return a data frame when the argument `data.table=FALSE` is used.

## Manual data entry

If you're a real masochist, use `edit`. You'll get a spreadsheet interface to enter numbers in.

```{r,eval=FALSE}
x<-data.frame()
x<-edit(x)
```

## Reading from Excel

Use the package `readxl`. It has two functions: `excel_sheets` and `read_excel`.

Choose an excel file using the dialog box and read it into R.

```{r,echo=TRUE,eval=FALSE}
library(readxl)
f1<-file.choose() #opens dialog box
d1<-excel_sheets(f1)
class(d1)
d2<-read_excel(f1) #default is first sheet
```

## Reading from Excel

This is the example from ?read_excel.
```{r}
suppressWarnings(library(readxl))
datasets <- system.file("extdata/datasets.xlsx", package = "readxl")
excel_sheets(datasets)
head(read_excel(datasets,sheet="mtcars"))

```

## Saving and loading R data sets.

When you close R you get the question "Save Workspace Image?" If you answer "Yes" it'll create a copy of your environment and save it in the file .RData. (This uses the `save.image` function.)

If you wish to save only certain objects the use `save` and provide a list of objects and a file name. Some people use .rda as the suffix for these files.
