---
title: "Loss Distributions"
output:
  slidy_presentation:
    duration: 75
---

## Loss Distributions
  
* actuar
* Direct optimization

## Fitting distributions
  
* Maximum likelihood
* Minimum (squared) distance
* optim
* q-q plots
* chi2 test

## actuar

```{r message=FALSE}
library(actuar)
data(dental)
```

* Contains the dental claims data from "Loss Models"
* `emm` calculates empirical moments
* `lev` limited expected value
* `coverage` modifies a loss distribution for coverage elements
* `fitdistr` will fit a distribution to a loss distribution function

## `fitdistr`

```{r }
library(MASS)
y <- fitdistr(dental, "gamma")
class(y)
```

## More loss data

```{r }
set.seed(8910)
years <- 2001:2010
frequency <- 1000

N <- rpois(length(years), frequency)

sevShape <- 2
sevScale <- 1000
severity <- rgamma(sum(N), sevShape, scale = sevScale)

fitGamma <- fitdistr(severity, "gamma")
fitLognormal <- fitdistr(severity, "lognormal")
```

## Compare fit to histogram

```{r }
sampleLogMean <- fitLognormal$estimate[1]
sampleLogSd <- fitLognormal$estimate[2]

sampleShape <- fitGamma$estimate[1]
sampleRate <- fitGamma$estimate[2]

x <- seq(0, max(severity), length.out=500)
yLN <- dlnorm(x, sampleLogMean, sampleLogSd)
yGamma <- dgamma(x, sampleShape, sampleRate)

hist(severity, freq=FALSE, ylim=range(yLN, yGamma))

lines(x, yLN, col="blue")

lines(x, yGamma, col="red")
```

## q-q plot

```{r }
probabilities = (1:(sum(N)))/(sum(N)+1)

gammaQ <- qgamma(probabilities, sampleShape, sampleRate)
plot(sort(gammaQ), sort(severity), xlab = 'Theoretical Quantiles', ylab = 'Sample Quantiles', pch=19)
abline(0,1)
```

## q-q plot

```{r }
lnQ <- qlnorm(probabilities, sampleLogMean, sampleLogSd)
plot(sort(lnQ), sort(severity), xlab = 'Theoretical Quantiles', ylab = 'Sample Quantiles', pch=19)
abline(0,1)
```

## Kolmogorov-Smirnov

```{r }
testGamma <- ks.test(severity, "pgamma", sampleShape, sampleRate)
testLN <- ks.test(severity, "plnorm", sampleLogMean, sampleLogSd)
testGamma2 <- ks.test(severity, "pgamma", sampleShape, 1)
```

## Useful resources

[http://cran.r-project.org/doc/contrib/Ricci-distributions-en.pdf](http://cran.r-project.org/doc/contrib/Ricci-distributions-en.pdf)

## Questions

* Draw a lognormal distribution with a mean of $10,000 and a CV of 30%.
* For that distribution, what is the probability of seeing a claim greater than $100,000?
* Generate 100 and 1,000 observations from that distribution.
* Draw a histogram for each sample.
* What are the mean, standard deviation and CV of each sample?
* Convince yourself that the sample data were not produced by a Weibull distribution.

## Answers

```{r }
severity <- 10000
CV <- .3
sigma <- sqrt(log(1 + CV^2))
mu <- log(severity) - sigma^2/2
plot(function(x) dlnorm(x), mu, sigma, ylab="LN f(x)")
```

##

```{r }
set.seed(1234)
claims = rlnorm(100, meanlog=log(30000), sdlog=1)
hist(claims, breaks=seq(1, 500000, length.out=40))
```

## Direct optimization

The `optim` function will optimize a function. Works very similar to the Solver algorithm in Excel. `optim` takes a function as an argument, so let's create a function.

```{r}
quadraticFun <- function(a, b, c){
  function(x) a*x^2 + b*x + c
}

myQuad <- quadraticFun(a=4, b=-3, c=3)
plot(myQuad, -10, 10)
```

## Direct optimization

8 is our initial guess. A good initial guess will speed up conversion.

```{r }
myResult <- optim(8, myQuad)
myResult
```

## Direct optimization
Default is to minimize. Set the parameter `fnscale` to something negative to convert to a maximization problem.

```{r }
myOtherQuad <- quadraticFun(-6, 20, -5)
plot(myOtherQuad, -10, 10)
myResult <- optim(8, myOtherQuad)
myResult <- optim(8, myOtherQuad, control = list(fnscale=-1))
```

## Direct optimization

Direct optimization allows us to create another objective function to maximize, or work with loss distributions for which there isn't yet support in a package like `actuar`. May be used for general purpose optimization problems, e.g. maximize rate of return for various capital allocation methods.

Note that optimization is a general, solved problem. Things like the simplex method already have package solutions in R. You don't need to reinvent the wheel!
