---
title: "Morning Recap"
author: "CAS RLAS 2015"
date: "December 14, 2015"
output: ioslides_presentation
---

## Morning Topics

Any Questions?

* Introductions
* Getting Started
* Vectors
* Data
* Loss Distributions
* Visualization

## Problem 1

Use the lognormal distribution.. Generate a large random sample. Do some of the following

* Create a vector of summary statistics.
* Create a histogram log of the distribution to see if it looks normal.
* Use `qnorm` as another check to see if the log is a normal distribution.

## Problem 2

Using the `Teams` datset in the `Lahman` package, choose a `teamID`.

* Create a data frame with `W`, `L`, `attendance` for years 1980 - 2014 for your selected team.
* Create a scatterplot to explore the relationship between wins and attendance.
* Determine the National League team that had the most wins in the 1920's.

## Problem 3

Using the iris dataset,

* Create different boxplots for the Sepal Length by species and display them in the same plot.
* Modify the plot by adding a descriptive title
* Plot Petal Width against Petal Length with points colored according to Species.
* Add a legend.

## Problem 1 Solution

The lognormal distribution has parameters `meanlog` and `sdlog`. This means that if X is lognormal then logX is normal with mean `meanlog` and standard deviation `sdlog`.

```{r}
  num<-10000;  ml<-5;  sdl<-2
  x1<-rlnorm(num,meanlog=ml,sdlog=sdl)
  q1<-quantile(x1,probs=seq(from=.1,to=.9,by=.1))
  mySummary<-c(min(x1),max(x1),mean(x1),sd(x1),q1)
  names(mySummary)[1:4]<-c("Min","Max","Mean","SD")
  mySummary
```

***

````{r}
y1<-log(x1)
(m1<-mean(y1))
(s1<-sd(y1))
```

***
```{r}
hist(y1,freq=FALSE)
```

***
```{r}
qqnorm(y1)
```


## Problem 2 Solution

```{r}
  suppressWarnings(library(Lahman)); data(Teams)
```

Filter the years and select the columns `yearID`, `teamID`, `W`, and `attendance`.
```{r}
  data1<-Teams[Teams$yearID>1979,c("yearID","teamID","W","attendance")]
  head(data1)
```

***
 Let's choose Atlanta.
 
```{r}
  atl<-data1[data1$teamID=="ATL",]
  plot(x=atl$W,y=atl$attendance,pch=19,xlab="Wins",ylab="Attendance")
  title(main="Atlanta \n 1980-2104")
```
 
***
```{r}
data1920<-Teams[Teams$yearID > 1919 & Teams$yearID<1930 
                & Teams$lgID=="NL",c("yearID","teamID","W")]
head(data1920)
```

***
```{r}
  totalWins<-aggregate(W~teamID,data=data1920,FUN=sum)
  dim(totalWins)
  totalWins
```

## Problem 3 Solution

```{r}
data(iris)
boxplot(Sepal.Length~Species,data=iris,xlab="Species")
title(main="Sepal Length of Iris Species")
```

***
```{r}
plot(x=iris$Petal.Length,y=iris$Petal.Width,pch=19,cex=.5,
     bty="l",xlab="Petal Length",ylab="Petal Width",
     col=as.numeric(iris$Species))
legend("topleft",fill=palette()[1:3],
       legend=levels(iris$Species),bty="n")
```

