---
title: "Reserving in R"
output:
  slidy_presentation:
    duration: 45
    fig_height: 6
    fig_width: 10
---

## Reserving in R

* ChainLadder
* MRMR

## ChainLadder

Originally developed by Markus Gesmann in late 2007. Now developed by MG with Daniel Murphy, Wayne Zhang, Alessandro Carrato, Giuseppe Crupi, Mario Wüthrich and Fabio Concina.

Implements many standard reserving techniques:
      
* Multiplicative chain ladder
* Munich chain ladder
* Bootstrap
* Dave Clark curve fitting    

## Basics
```{r message=FALSE}
library(ChainLadder, verbose=FALSE)
data(RAA)

typeof(RAA)
is.matrix(RAA)
```

## Have a look at the data
```{r }
RAA
```


## Plot the data
```{r fig.height=4.5}
plot(RAA)
```

## Plot option lattice

```{r }
plot(RAA, lattice=TRUE)
```

## Calculation
```{r }
results = MackChainLadder(Triangle = RAA, est.sigma="Mack")

typeof(results)
class(results)
```

## Plot the results

```{r }
plot(results)
```

## Plot the results as a lattice

```{r }
plot(results, lattice=TRUE)
```

## Munich chain ladder
```{r warning=FALSE, message=FALSE}
data(MCLpaid)
data(MCLincurred)

is.matrix(MCLpaid)

MunichChainLadder(MCLpaid, MCLincurred, 
 est.sigmaP = "log-linear", 
 est.sigmaI = "log-linear", 
 tailP=FALSE, tailI=FALSE) 
```

## Werner-Modlin data

Data taken from Appendix A of Werner/Modlin.
[http://www.casact.org/library/studynotes/werner_modlin_ratemaking.pdf].

```{r }
library(raw)
data(PPA)
head(PPA_LossDevelopment)
```

## Convert to ChainLadder triangle
```{r }
triPPA <- ChainLadder::as.triangle(PPA_LossDevelopment
                                   , origin="AccidentYear"
                                   , dev="Lag"
                                   , value="ReportedLossPaidALAE")
head(triPPA)
```

## Clark curve

```{r }
myClarkFit <- ClarkLDF(triPPA, maxage=Inf)
```

## Clark LDF

```{r }
plot(myClarkFit)
```

