---
title: "Advanced Visualization"
output: slidy_presentation
---

## Topics

* ggplot2
* Maps

## ggplot2

ggplot2 developed by Hadley Wickham, based on the "grammar of graphics"
  
Particularly well suited for multi-dimensional, multivariate analysis.

Requires 3 things:

1. Data
2. Aesthetics
3. Layers

## Data

```{r }
library(raw)
data(RegionExperience)
library(ggplot2)

basePlot <- ggplot(RegionExperience)
class(basePlot)
```

## Aesthetics

Aesthetics are anything visible on the plot. Added with the `aes` function. Data will be mapped to visual elements.

```{r }
basePlot <- basePlot + aes(x = PolicyYear, y = NumClaims, group=Region, color=Region, fill=Region)
```

## Adding layers

The `geom_*` functions add geometric shapes. 

```{r }
p <- basePlot + geom_line()
p
```

## Adding layers

```{r }
p <- basePlot + geom_point()
p
```

## Nothing wrong with adding two layers

```{r }
p <- basePlot + geom_point() + geom_line()
p
```

## Adding layers

```{r }
p <- basePlot + geom_bar(stat="identity")
p
```

Here `stat` is a statistical transformation.

## One step

Typically we don't do this in steps.

```{r }
p <- ggplot(RegionExperience, aes(x = PolicyYear, y = NumClaims, group=Region, color=Region)) + geom_line()
p
```

## Facets
```{r }
p <- basePlot + geom_line() + facet_wrap(~ Region)
p
```

## Facets

That plot was a bit boring.

```{r }
data(StateExperience)
p = ggplot(StateExperience, aes(x = PolicyYear, y = NumClaims, color=Fullname)) + geom_point() + facet_wrap(~ Region)
p
```

## 
```{r eval=FALSE}
p <- ggplot(subset(StateExperience, Region == "South"), aes(x = PolicyYear, y = NumClaims, label=Postal))
p <- p + geom_point() + facet_wrap(~ Postal)
p
```

## Maps!
```{r }
library(maps)
map('state')
```

## Hurricanes
```{r }
data(Hurricane)
 
dfKatrina = subset(Hurricane, Name == 'KATRINA')
dfKatrina = dfKatrina[dfKatrina$Year == max(dfKatrina$Year), ]

 
dfHugo = subset(Hurricane, Name == 'HUGO')
dfHugo = dfHugo[dfHugo$Year == max(dfHugo$Year), ]
 
dfDonna = Hurricane[Hurricane$Name == 'DONNA', ]
dfDonna = dfDonna[dfDonna$Year == max(dfDonna$Year), ]
```

## Hurricanes
```{r }
map('state')
points(dfKatrina$Longitude, dfKatrina$Latitude, pch=19, col = 'red')
points(dfHugo$Longitude, dfHugo$Latitude, pch = 19, col = 'blue')
points(dfDonna$Longitude, dfDonna$Latitude, pch = 19, col = 'green')
```

## Choropleths
