---
title: "Advanced Topics"
author: "R LAS"
date: "December 15"
output: ioslides_presentation
---

# Miscellaneous stuff

## `assign`, `eval`, and `parse`

`assign` allows you to assign a value to a character string.

```{r}
assign("x",1)
x
```

This was used in the programming section. It's helpful when you don't want to make the code clearer.

```{r}
x<--3
x
```
***

`eval` evaluates an expression. Use `parse` with `eval` to evaluate a character string.

```{r}
x<-1
eval(x)
eval("x")

eval(parse(text="x"))
```

## Reading data off the internet

```{r}
library(XML)
nflStandings<-"http://sports.yahoo.com/nfl/standings/"
nflTables <- readHTMLTable(nflStandings)

class(nflTables)
```

***
```{r}
length(nflTables)
sapply(nflTables,class)
```

***
```{r}
head(lapply(nflTables,dim))
```

***
It looks like the first one is different, let's check.

```{r}
str(nflTables[[1]])
str(nflTables[[2]])
```

***
Look at the first one with data.
```{r}
nflTables[[2]][,1:7]
```

***
Lets add the conference and division names to the list and turn it into a data frame.

```{r}
division<-names(nflTables)[2:9]
conference<-c(rep("AFC",4),rep("NFC",5))

nflTables<-lapply(2:9,function(i) 
  cbind(conference[i],division[i],nflTables[[i]][,1:7]))

length(nflTables)
```

***
```{r}
nflTables[[1]]
```

***
Use `do.call` to bind an unknown number of data frames.
```{r}
nflStandings<-do.call(rbind,nflTables)
names(nflStandings)[1:2]<-c("Conference","Division")
dim(nflStandings)
head(nflStandings)
```

## Colors

R has `r length(colors())` different colors. Additional information is [here.](http://research.stowers-institute.org/efg/R/Color/Chart/) 

```{r}
palette()
head(colors())
```

## Ten Random Colors
```{r}
set.seed(6345789)
numColors<-10
n1<-sample(1:length(colors()),replace=FALSE,
           size=numColors)
pie(rep(1,numColors),col=colors()[n1],
    labels=colors()[n1])
title(main="A Color Wheel")
```

## The default palette
```{r}
pie(rep(1,length(palette())),col=palette(),
    labels=palette())
title(main="Default")
```


## The rainbow pallette

```{r,fig.width=3,fig.height=3}
numColors<-8
pie(rep(1,numColors),col=rainbow(numColors),
    labels=rainbow(numColors))
title(main=paste("rainbow(",numColors,")",sep=""))
title(sub="Note that the names are in RGB")
```

##R Color Brewer

An overview of color schemes is at [colorbrewer2.org](colorbrewer2.org). They're implemented in the RColorBrewer package.

```{r,fig.width=4,fig.height=3}
library(RColorBrewer)
display.brewer.pal(6,"RdBu")
```

# Dplyr

## Overview

[Dplyr](http://blog.rstudio.org/2015/09/04/dplyr-0-4-3/) is another RStudio package that is designed specifically to handle the tasks covered in the data frames presentation.

* select - pick columns by name
* filter - keep rows matching criteria
* mutate - add new variables
* arrange - reorder rows
* summarize - reduce variables to values

A major advantage of `dplyr` is the speed at which it handles large data sets.An alternative to `dplyr` is the `data.table` package.

There's also a `dplyr` [cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) 

***

We'll only cover the basics as there are a number of additional resources available.

### Hadley Wickham Presentations
* [RStudio Webinar: The Grammar and Graphics of Data Science](https://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html)
* [Hadley Wickham's Dplyr Tutorial](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) 
* [UseR! 2014 Presentation Part 1](https://www.youtube.com/watch?v=8SGif63VW6E)
* [UseR! 2014 Presentation Part 2](https://www.youtube.com/watch?v=Ue08LVuk790)

### Others

* [Data School's Tutorial](https://www.youtube.com/watch?v=jWjqLW-u3hc&feature=youtu.be)



## NYC Flight Data

The above referenced tutorials and presntations use data from the `nycflights13` package. On-time data for all flights that departed NYC (JFK, LGA, or EWR) in 2013.

```{r,message=FALSE}
library(dplyr)
library(nycflights13)
data(flights)
```

***
```{r}
class(flights)
dim(flights)
```

Note that flights is of class `tbl_df`. The main advantage of this class is it only prints what can fit on the screen and `[` doesn't need `drop=FALSE`.

***
```{r}
flights
```

## Lahman Database 
Since the flights data is used in the other resources, we'll continue using the Lahman baseball database that we used earlier.

```{r}
library(Lahman)
class(Master)
```


## Filter

Find Babe Ruth in the Master player table. It'll be a subset of a data frame so it's a data frame. We need to convert `Master` to a `tbl_df` (this isn't necessary if you've used `read_csv` in the `readr` package.)
```{r}
names(Master)
Master<-tbl_df(Master)
ruthTable<-filter(Master,nameLast=="Ruth")
```

***
```{r}
class(ruthTable)
ruthTable
```

***

Observations:

* The first argument in `filter` is the data name.
* We used the column name within the function without referring to the data.
* A `tbl_df` object was returned.

***

Suppose we did the same thing with Ted Williams. There are more than one player with a last name of Williams.

```{r}
williamsTable<-filter(Master,nameLast=="Williams")
nrow(williamsTable)
```

```{r}
tedWilliams<-filter(williamsTable,nameFirst=="Ted")
tedWilliams
```

***
Of course, we could have created the table with one command.
```{r}
tw<-filter(Master,nameFirst=="Ted" & nameLast=="Williams")
tw
```

## Select
Now lets select the `playerID`, `nameLast`, `nameFirst`,`bats`, and `throws`.
```{r}
select(Master,playerID,nameLast,nameFirst,bats,throws)
```

## Combine `select` and `filter`

The pipe operator `%>%` allows for a syntax that mirrors your thought process.

* Start with the data frame then (`%>%`)
* Convert it to an object of class `tbl_df` then (`%>%`)
* `select` the columns you want then (`%>%`)
* `filter` by the desired criteria.

***

```{r} 
Master %>% tbl_df() %>%
      select(playerID,nameLast,nameFirst,bats,throws) %>%
      filter(nameLast=="Williams" & nameFirst=="Ted")
```

## Mutate

Use the batting table to calculate a batting average (H/AB). Use only players with at least 600 AB from 2000-2014. (Note that you'll only know the player's ID and not his name.)

```{r}
tempBatting <- Batting %>% tbl_df %>% filter(yearID > 1999 & AB >599) %>%
  select(playerID,yearID,AB,H) %>% mutate(AVG=round(H/AB,3))
tempBatting
```

## Arrange

Add the Player Info and sort in descending order.

```{r}

inner_join(tempBatting,Master,by="playerID") %>%
  select(nameLast,nameFirst,yearID,AB,H,AVG) %>%
  arrange(desc(AVG))
```

## Summarize

* Pitchers with no games started in the 1960's
* Left-handed.
* Played at least 5 seasons
* Compute strikeouts per 27 outs

```{r}
Master<-Master %>% tbl_df()
Pitching<-Pitching %>% tbl_df()
temp<-inner_join(Pitching,Master,by="playerID") %>%
 filter(yearID >1959 & yearID <=1969) %>%
  filter(throws=="L" & GS==0) %>%
  select(playerID,nameLast,nameFirst,G,IPouts,SO) %>%
  group_by(playerID,nameLast,nameFirst) %>% summarize(Seasons=n(),Games=sum(G),Outs=sum(IPouts),SO=sum(SO)) %>%
  mutate(SO_27=round((SO/Outs)*27,1)) %>%
  filter(Seasons>=5) %>% 
  select(nameLast,nameFirst,Seasons,Games,Outs,SO,SO_27)
```

***
```{r}
temp
```


  