---
title: "Vectors"
output:
  slidy_presentation:
    duration: 45
    fig_height: 3
    fig_width: 6
  beamer_presentation:
    fonttheme: professionalfonts
---

## Introducing vectors

In R, every variable is a vector. Think of a set of contiguous cells in a spreadsheet. 

```{r eval=TRUE}
set.seed(1234)
e  <- rnorm(100)
X1 <- 1:10
```

Here, `e` is a vector with `N` values. `X1` is the sequence of integers from 1 through 10. 

Vectors can grow and shrink automatically. No need to move cells around on a sheet. No need to copy formulas or change named ranges.

## Vector properties

* Every element in a vector must be the same type. 
    * R will change data types if they are not!
    * Different types are possible by using a list or a data frame (later)
* Vectors have one dimension
    * Higher dimensions are possible via matrices and arrays
* Possible to add metadata (like names) via attributes

## Vector construction

Vectors are constructed in one of several ways:

* Return from a function or operation
    * `seq`, `rep`, `sample`, `rnorm`,  etc.
* Concatenation
* Growth by assignment

## `seq`

`seq` is used often to generate a sequence of values. The colon operator `:` is a shortcut for a sequence of integers.

```{r }
pies = seq(from = 0, by = pi, length.out = 5)
i <- 1:5
year = 2000:2004
```

```{r echo=FALSE}
library(rblocks)
i <- 1:5
block = make_block(i, type = 'vector')
block
```

## `rep`

The `rep` function will replicate its input 
```{r }
i = rep(pi, 100)
head(i)
```

## Concatenation

The `c()` function will concatenate values.

```{r results='hide'}
i <- c(1, 2, 3, 4, 5)
j <- c(6, 7, 8, 9, 10)
k <- c(i, j)
l <- c(1:5, 6:10)
```

```{r echo=FALSE}
i = c(1, 2, 3, 4, 5)
j = c(6, 7, 8, 9, 10)
k = c(i, j)
block_i = make_block(i, type = 'vector')
block_j = make_block(j, type = 'vector')
block_i[1:5] = "red"
block_j[1:5] = "blue"
block_k = make_block(c(block_i,block_j), type='vector')
block_k[1:5] = "red"
block_k[6:10] <- "blue"

block_i
block_j


block_k
```

## Growth by assignment

Assigning a value beyond a vectors limits will automatically grow the vector. Interim values are assigned `NA`.

```{r }
i <- 1:10
i[30] = pi
i
```

## Vector access - by index

Vectors may be accessed by their numeric indices. Remember, ':' is shorthand to generate a sequence.

```{r }
set.seed(1234)
e <- rnorm(100)
e[1]
e[1:4]
e[c(1,3)]
```

## Vector access - logical access

Vectors may be accessed logically. This may be done by passing in a logical vector, or a logical expression.

```{r }
i = 5:9
i[c(TRUE, FALSE, FALSE, FALSE, TRUE)]
i[i > 7]
b = i > 7
b
i[b]
```

## `which`

The `which` function returns indices that match a logical expression.

```{r }
i <- 11:20
which(i > 12)
i[which(i > 12)]
```

## `sample`

The `sample` function will generate a random sample. Great to use for randomizing a vector.

```{r }
months <- c("January", "February", "March", "April"
            , "May", "June", "July", "August"
            , "September", "October", "November", "December")

set.seed(1234)
mixedMonths <- sample(months)
head(mixedMonths)
```

Get lots of months with the `size` parameter:
```{r }
set.seed(1234)
lotsOfMonths <- sample(months, size = 100, replace = TRUE)
head(lotsOfMonths)
```

## `sample` II

Sample may also be used within the indexing of the vector itself:

```{r }
set.seed(1234)
moreMonths <- months[sample(1:12, replace=TRUE, size=100)]
head(moreMonths)

# Cleaner with sample.int
set.seed(1234)
evenMoreMonths <- months[sample.int(length(months), size=100, replace=TRUE)]
head(evenMoreMonths)
```

## `order`

The function `order` will return the indices of the vector in order.

```{r }
set.seed(1234)
x <- sample(1:10)
x
order(x)
x[order(x)]
```


## Vector arithmetic

Vectors may be used in arithmetic operations. 
  
```{r eval=FALSE}
B0 <- 5
B1 <- 1.5

set.seed(1234)

e <- rnorm(N, mean = 0, sd = 1)
X1 <- rep(seq(1,10),10)

Y <- B0 + B1 * X1 + e
```

Y is now a vector with length equal to the longest vector used in the calculation.

Question: B0 and B1 are vectors of length 1. 

X1 and e are vectors of length 100. 

How are they combined?


## Recycling

R will "recycle" vectors until there are enough elements to perform an operation. Everything gets as "long" as the longest vector in the operation. For scalar operations on a vector this doesn't involve any drama. Try the following code:

```{r size='tiny'}
vector1 = 1:10
vector2 = 1:5
scalar = 3

print(vector1 + scalar)
print(vector2 + scalar)
print(vector1 + vector2)
```

## Set theory - Part I

The `%in%` operator will return a logical vector indicating whether or not an element of the first set is contained in the second set.

```{r }
x <- 1:10
y <- 5:15
x %in% y
```

## Set theory - Part II

* `union`
* `intersect`
* `setdiff`
* `setequal`
* `is.element`

```{r eval = FALSE}
?union
```

```{r }
x <- 1900:1910
y <- 1905:1915
intersect(x, y)
setdiff(x, y)
setequal(x, y)
is.element(1941, y)
```

## Summarization

Loads of functions take vector input and return scalar output. Translation of a large sest of numbers into a few, informative values is one of the cornerstones of statistics.

```{r eval=FALSE}
x = 1:50
sum(x)
mean(x)
max(x)
length(x)
var(x)
```

## Vectors 

Vectors are like atoms. If you understand vectors- how to create them, how to manipulate them, how to access the elements, you're well on your way to grasping how to handle other objects in R. 

Vectors may combine to form molecules, but fundamentally, _everything_ in R is a vector.

## Exercise - Vectors

Create a vector of length 10, with years starting from 1980.

Create a vector with values from 1972 to 2012 in increments of four (1972, 1976, 1980, etc.)

Construct the following vectors (feel free to use the `VectorQuestion.R` script):
```{r }
FirstName <- c("Richard", "James", "Ronald", "Ronald"
              , "George", "William", "William", "George"
              , "George", "Barack", "Barack")
LastName <- c("Nixon", "Carter", "Reagan", "Reagan"
              , "Bush", "Clinton", "Clinton", "Bush"
              , "Bush", "Obama", "Obama")
ElectionYear <- seq(1972, 2012, 4)
```

* List the last names in alphabetical order
* List the years in order by first name.
* Create a vector of years when someone named "George" was elected.
* How many Georges were elected before 1996?
* Generate a random sample of 100 presidents.

## Answer
```{r }
LastName[order(LastName)]
ElectionYear[order(FirstName)]
ElectionYear[FirstName == 'George']
myLogical <- (FirstName == 'George') & (ElectionYear < 1996)
length(which(myLogical))
sum(myLogical)

sample(LastName, 100, replace = TRUE)
```