---
title: "Data"
output: ioslides_presentation
---

## Reading data

```{r eval=FALSE}
myData = read.csv("SomeFile.csv")
```

## Reading from Excel

Actually there are several ways:
* XLConnect
* 

```{r eval=FALSE}
library(XLConnect)
wbk = loadWorkbook("myWorkbook.xlsx")
df = readWorksheet(wbk, someSheet)
```

## Reading from the web - 1 

```{r eval=FALSE}
URL = "http://www.casact.org/research/reserve_data/ppauto_pos.csv"
df = read.csv(URL, stringsAsFactors = FALSE)
```

## Reading from the web - 2 

```{r eval=FALSE}
library(XML)
URL = "http://www.pro-football-reference.com/teams/nyj/2012_games.htm"
games = readHTMLTable(URL, stringsAsFactors = FALSE)
```

## Reading from a database 

```{r eval=FALSE}
library(RODBC)
myChannel = odbcConnect(dsn = "MyDSN_Name")
df = sqlQuery(myChannel, "SELECT stuff FROM myTable")
```

## Data frames 
Every time we read in data, we're creating a data frame. The data frame is a seminal concept in R. Most statistical operations expect one and they are the most common way to pass data in and out of R.

Although critical to understand, this is very, very easy to get. What's a data frame? It's a table. That's it.

Much more on data frames in the next session.

## Read some data
```{r eval=FALSE}
df = read.csv("../Data/StateData.csv")
```

```{r eval=FALSE}
View(df)
```

---

Can everyone load the State data into a data frame?
