---
title: "Basic programming concepts"
output: ioslides_presentation
---

## Writing your own functions
```{r}
myFunction <- function(x, y){
  x + y
```

## A real function
```{r}
CircleArea <- function(radius){
  area <- (2 * pi * radius) ^ 2
  area
}
```

---

Functions are basically the same as functions entered into a cell in a spreadsheet. Functions may call other functions and so on.

Exercise: Write a function to compute the roots of a quadratic equation.

## Answer

```{r}
QuadRoot <- function(a, b, c){
  disc <- b^2 - 4 * a * c
  result <- -2 * b - sqrt(disc)
  result <- result / (2 * a)
  result
```

## Flow control - if

```{r }
if (something) {
  
} else {
  
}
```

## Flow control - for
```{r }
for (i in 1:5) {
  
}
```

## Flow control - while
```{r}
while (something) {
  
}
```

## Logical operators

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Equality | List          |
| 2d | inequality      | Data frame    |
| nd | And    |               |
       Or
       
