---
title: "Basic programming concepts"
output:
  slidy_presentation:
    duration: 30
---

## Part 5 - Basic programming

* Writing your own functions
* Control structures

## Writing your own functions
```{r}
myFunction <- function(x, y){
  x + y
}
```

## A real function
```{r}
CircleArea <- function(radius){
  area <- (2 * pi * radius) ^ 2
  area
}
```

---

Functions are basically the same as functions entered into a cell in a spreadsheet. Functions may call other functions and so on.

Exercise: Write a function to compute the roots of a quadratic equation.

## Answer

```{r}
QuadRoot <- function(a, b, c){
  disc <- b^2 - 4 * a * c
  result <- -2 * b - sqrt(disc)
  result <- result / (2 * a)
  result
}
```

## Flow control - if

```{r eval=FALSE}
if (something) {
  
} else {
  
}
```

## Logical operators

## 
```{r }
x <- 1:10
range(x)

seq_along(x)
seq_len(100)
```
       
## Flow control - for
```{r eval=FALSE}
for (i in 1:5) {
  print(i)
}

for (i in seq_len(5)){ print(letters[i]) }
```

## Flow control - while
```{r eval=FALSE}
while (x != 'q') {
  
}
```

##

Flow control in R works similarly to other languages. However, you don't need it as often.

Compare:
```{r }
myIndex <- 1
while (letters[myIndex] != 'q') {
  myIndex <- myIndex + 1
}

myOtherIndex <- which(letters == 'q')

identical(myIndex, myOtherIndex)
all.equal(myIndex, myOtherIndex)
```

## 

Flow control is needed when:

* Recursive operation
* Optimization

*NOT* needed when:
* Applying the same operation to many elements
* Searching
* Summarizing

## Questions

* Write a function that will compute the sum of squared error between two vectors.
* Write a function that will return the number of elements of a vector needed to reach a sum of 500.
* Write a function that will minimize the sum of squared error between two vectors by applying a linear transform to the first of the two vectors. The function will return the linear transform parameters as a list with two elements.

## Answers

```{r }
```
