---
title: "File commands"
author: "Brian A. Fannin"
date: "11/14/2014"
output: 
  pdf_document:
    number_sections: yes
---

## Typical operations

* Determine if a file exists
* Get basic information about a file
* Delete a file or directory
* Copy a file
* Move a file
* Create a new file

## OS differences:

* Windows favors the backslash: "/"
* *nix(which includes MacOS) use forward slash: "\"
* When using 'file.exists' and the like, cannot put a trailing slash under Windows.
* Unix/Linux may use 'symlinks'.

## Constructing the filename

file.path works faster than paste. Also takes care of OS differences.
```{r }
file.path()
```

## Determine if a file exists

Returns a logical vector to May use wildcards or regular expressions.
```{r }
```

```

A 'file' may also mean a directory. 

## file.info

Returns information as a data frame.

| Column | Type    | 
|--------|---------|
| size   | double  |
| isdir  | logical |
| mode   | integer |
| mtime  | POSIXct |
| ctime  | POSIXct |
| atime  | POSIXct |
| uid    | integer |
| gid    | integer |
| uname  | uname   |
| grname | grname  |

## 

getwd()
setwd()
list.files()
file.info()
copy
delete
create a directory

## file.create

```{r eval=FALSE}
myNewFilename = "MyFile.txt"
if(!file.exists(myNewFilename)) {
  # Very dangerous! 
  unlink(myNewFilename)
  }

file.create("MyNewFile.txt")
```

This document took inspiration from "Advanced R" by Hadley Wickham.