---
title: "Untitled"
author: "Brian A. Fannin"
date: "11/28/2014"
output: slidy_presentation
---

## Advanced Modeling

* GLMs
* Basic machine learning

## Logistic regression

* Maps the real numbers to the space between 0 and 1.
* Binomial parameter `p` is a function.
* Will a claim close?
* Will a claimant use an attorney?
* Will a policyholder renew at a particular premium?

## Logistic Regression - an example
```{r }
library(raw)
data(NFL)
NFL = subset(NFL, Home==TRUE)
```

## Logistic Regression
```{r }
plot(jitter(NFL$TODiff), jitter(NFL$Win, factor=0.5), xlab="Turnover difference", ylab = "Wins")
```

## Logistic Regression

Null deviance may be used to compare competing models. RMSE, SSE may be used to assess how well a model works. Also, compare results for test and training data sets.

```{r }
fit = glm(Win ~ TODiff, family=binomial(link="logit"), data=NFL)
summary(fit)
```

## Logistic Regression
```{r }
library(boot)
plot(jitter(NFL$TODiff), jitter(NFL$Win, factor=0.5), xlab="Turnover difference", ylab = "Wins")
curve(inv.logit(coef(fit)[1] + coef(fit)[2]*x), add=TRUE)
abline(h=0.5)
```

## Machine learning

* Clustering
* 