---
title: "Reserving in R"
output:
  slidy_presentation:
    duration: 75
    fig_height: 6
    fig_width: 10
---

## Reserving in R

* ChainLadder
* MRMR

## ChainLadder

Originally developed by Markus Gesmann in late 2007. Now developed by MG with Dan Murphy and Wayne Zhang

Implements many standard reserving techniques:
      
* Multiplicative chain ladder
* Munich chain ladder
* Bootstrap
* Dave Clark curve fitting    

## Basics
```{r message=FALSE}
library(ChainLadder, verbose=FALSE)
data(RAA)

typeof(RAA)
is.matrix(RAA)
```

## Have a look at the data
```{r }
RAA
```


## Plot the data
```{r fig.height=4.5}
plot(RAA)
```

## Plot option lattice

```{r }
plot(RAA, lattice=TRUE)
```

## Calculation
```{r }
results = MackChainLadder(Triangle = RAA, est.sigma="Mack")

typeof(results)
class(results)
```

## Plot the results

```{r }
plot(results)
```

## Plot the results as a lattice

```{r }
plot(results, lattice=TRUE)
```

## Munich chain ladder
```{r warning=FALSE, message=FALSE}
data(MCLpaid)
data(MCLincurred)

is.matrix(MCLpaid)

MunichChainLadder(MCLpaid, MCLincurred, 
 est.sigmaP = "log-linear", 
 est.sigmaI = "log-linear", 
 tailP=FALSE, tailI=FALSE) 
```

## Werner-Modlin data

Data taken from Appendix A of Werner/Modlin.
[http://www.casact.org/library/studynotes/werner_modlin_ratemaking.pdf].

```{r }
library(raw)
data(PPA)
head(PPA_LossDevelopment)
```

## Convert to ChainLadder triangle
```{r }
triPPA <- ChainLadder::as.triangle(PPA_LossDevelopment
                                   , origin="AccidentYear"
                                   , dev="Lag"
                                   , value="ReportedLossPaidALAE")
head(triPPA)
```

## Clark curve

```{r }
myClarkFit <- ClarkLDF(triPPA, maxage=Inf)
```

## Clark LDF

```{r }
plot(myClarkFit)
```

## What is MRMR?

- Flexible framework for reserving based on linear models
- Linear models = Chain ladder and Additive methods
- Emphasis on:
    - Ease of use
    - Multi-dimensional treatment of data
    - Visualization supports exploratory analysis and assessment of model fit

## Inspirations

- Leigh Halliwell, Jim Guszcza
- Gelman & Hill, "Data Analysis Using Regression and Multilevel/Hierarchical Models"
- ggplot2, lubridate, sp

## Object structure

- OriginPeriod
- St*Measures
- Triangle
- TriangleModel
- TriangleProjection

## Common object properties

- May be thought of as data frames with constraints and metadata
- `[`, `$` accessors
- `rbind`, `c`, `length`
- New generics: 
    - `write.excel`
    - `Grow`

## OriginPeriod

```{r results='hide'}
library(MRMR)
library(lubridate)
data(NAIC)
op = OriginPeriod(StartDate = as.Date("2001-01-01")
                  , NumPeriods = 10
                  , Period=as.period(1, "years")
                  , Type="Accident Year")
op$Moniker = paste0("AY ", year(op$StartDate))
```

## OriginPeriod (cont'd)

```{r echo=TRUE, results='hide'}
head(as.data.frame(op)[, 1:3])
```
```{r echo=FALSE, results='asis', size='huge', message=FALSE, tidy=FALSE, highlight=FALSE}
library(pander)
panderOptions("table.style", "rmarkdown")
dfDisplay = as.data.frame(op)
pander(head(dfDisplay))
```

## Non-standard OriginPeriods

```{r echo=TRUE, results='hide'}
startDates = seq(as.Date("2001/01/01")
                 , as.Date("2005/12/31"), by="6 months")
endDates = startDates + as.period(6, "months") - days(1)
op = OriginPeriod(startDates, endDates)
```
```{r echo=FALSE, results='asis', size='huge', message=FALSE, tidy=FALSE, highlight=FALSE}
panderOptions("table.style", "rmarkdown")
dfDisplay = as.data.frame(op)
pander(head(dfDisplay))
```

## Member access
```{r echo=TRUE, results='hide'}
op = OriginPeriod(seq(2001:2010))
op$Moniker = paste0("AY ", year(op$StartDate))
x = op[1]
y = op[2:3]
z = c(x, y)
y = op[c("AY 2004", "AY 2008")]
y = op$StartDate
y = op$Type
y = op$Moniker[3]
```

## St*Measures

- St*ticMeasures contain a set of observations associated with an OriginPeriod
- StaticMeasure objects do not vary over time
    + Earned Premium, Payroll, Number of autos
- StochasticMeasure objects vary over time
    + Paid loss, number of open claims

## St*Measures (cont'd)

- Contain a list of Levels of arbitray complexity
    + Each Level has a name and a set of attributes
    + Ex: Line level may have attributes "GL", "Motor", etc.
- Contain a set of Measures
    + StochasticMeasure will automatically compute incremental and prior measures

## Examples

All examples use NAIC data compiled by Glen Meyers and Peng Shi of ISO. Available at [http://www.casact.org/research/index.cfm?fa=loss_reserves_data](http://www.casact.org/research/index.cfm?fa=loss_reserves_data)

## StaticMeasure construction

```{r results='hide'}
op = OriginPeriod(StartDate = as.Date("1988-01-01")
                  , NumPeriods = 10
                  , Period=as.period(1, "year")
                  , Type="Accident Year")
companies <- unique(dfWC$Company)
smWC = StaticMeasure(OriginPeriod = op
                     , Level = list(Company=companies, Line="WC")
                     , Measure=c("DirectEP", "NetEP")
                     , Data=dfWC[dfWC$Lag == 1, ])
```

## StaticMeasure plot

```{r fig.height=6}
plot(smWC)
```

## StaticMeasure facet plot

```{r fig.height=6}
plot(smWC, FacetFormula=as.formula(Line+Company~Measure))
```

## StaticMeasure multidimensional plot

```{r fig.height=6}
smPPA = StaticMeasure(OriginPeriod = op
                     , Level = list(Company=companies, Line="PPA")
                , Measure=c("DirectEP", "NetEP")
                , Data=dfPPA[dfPPA$Lag == 1, ])
smMulti = c(smPPA, smWC)
plot(smMulti, FacetFormula=as.formula(Company~Line))
```

## StochasticMeasure - Construction

```{r }
scmWC = StochasticMeasure(OriginPeriod = op
                        , Level=list(Company=companies, Line="WC")
                        , Measure = c("CumulativeIncurred", "CumulativePaid")
                        , DevPeriod = as.period(1, "year")
                        , Lags=1:10
                        , Data=dfWC
                        , OriginPeriodSort = "AccidentYear"
                        , EvaluationDates=seq.Date(as.Date("1988-12-31"), as.Date("2006-12-31"), by="1 year"))
```

## StochasticMeasure - 1

```{r fig.height=6}
plot(UpperTriangle(scmWC), Measure="CumulativeIncurred")
```

## StochasticMeasure - 2

```{r fig.height=6}
plot(UpperTriangle(scmWC), Measure="CumulativeIncurred", TimeAxis="EvaluationDate")
```

## Triangle

Very simple object. Composed of a StaticMeasure and a StochasticMeasure

```{r fig.height=6}
triWC = Triangle(smWC, scmWC, "Workers Comp Triangle")

plot(UpperTriangle(triWC)
     , Response="IncrementalPaid"
     , Predictor="PriorCumulativePaid")
```

## Triangle plot switch variables

```{r fig.height=6}
plot(UpperTriangle(triWC)
     , Response="IncrementalPaid"
     , Predictor="NetEP")
```

## TriangleModel

3 Model Types

- individual - Ignore the other data
- pooled - All data is equal
- blended - Weighted average of first two

## TriangleModel

```{r }
PaidCLPooled = 
  TriangleModel(UpperTriangle(triWC)
                , Response = "IncrementalPaid"
                , Predictor = "PriorCumulativePaid"
                , ModelType = "pooled")
```

## Residual plots
```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
PaidCLPooled = TriangleModel(UpperTriangle(triWC)
                             , Response = "IncrementalPaid"
                             , Predictor = "PriorCumulativePaid"
                             , ModelType = "pooled")

PaidCLIndividual = TriangleModel(UpperTriangle(triWC)
                             , Response = "IncrementalPaid"
                             , Predictor = "PriorCumulativePaid"
                             , ModelType = "individual")

PaidCLBlended = TriangleModel(UpperTriangle(triWC)
                              , Response = "IncrementalPaid"
                              , Predictor = "PriorCumulativePaid"
                              , Group = "Company"
                              , ModelType = "blended")
```
```{r warning=FALSE, fig.height=6}
PlotResiduals(PaidCLPooled)
```

## Residual plots 2

```{r warning=FALSE, fig.height=6}
PlotResiduals(PaidCLIndividual)
```

## Residual plots 3

```{r warning=FALSE, fig.height=6}
PlotResiduals(PaidCLBlended)
```

## Model comparison

```{r }
RMSE(PaidCLPooled)
RMSE(PaidCLIndividual)
RMSE(PaidCLBlended)
```

## Projected results

```{r }
projPaidCLPooled = 
  TriangleProjection(PaidCLPooled
                     , AsOfDate=as.Date("2006-12-31")
                     , MaxLag=10)
```

## Projected results

```{r echo=FALSE, results='asis'}
projPaidCLIndividual = TriangleProjection(PaidCLIndividual
                                      , AsOfDate=as.Date("2006-12-31")
                                      , MaxLag=10)

projPaidCLBlended = TriangleProjection(PaidCLBlended
                                      , AsOfDate=as.Date("2006-12-31")
                                      , MaxLag=10)

df.proj1 = as.data.frame(projPaidCLPooled@Projection)
df.proj2 = as.data.frame(projPaidCLIndividual@Projection)
df.proj3 = as.data.frame(projPaidCLBlended@Projection)

df.actual = as.data.frame(triWC)
companies = unique(df.actual$Company)
df.actual = subset(df.actual, Lag == 10)

df.actual = df.actual[, c("EvaluationDate", "IncrementalPaid")]
row.names(df.actual) = NULL

df.proj1 = df.proj1[, "IncrementalPaid"]
df.proj2 = df.proj2[, "IncrementalPaid"]
df.proj3 = df.proj3[, "IncrementalPaid"]

dfCompare = cbind(df.actual, df.proj1, df.proj2, df.proj3)
names(dfCompare)[2:5] = c("Actual", "Pooled", "Individual", "Blended")
dfCompare = subset(dfCompare, EvaluationDate > as.Date("1997-12-31"))

# panderOptions("table.style", "rmarkdown")
# pander(dfCompare[1:10, ])
```

## Which package?

ChainLadder:

* Excellent for learning/applying basic (and not-so-basic) techniques
* Great to explore which models may fit your data
* Very mature, stable package

MRMR

* Multivariate visualization
* Combine experience of individual groups
* Loss reserving with credibility
* Gateway drug to Bayesian reserving
* Probably harder to use if you're just starting out with R
