---
title: "Visualization"
output:
  slidy_presentation:
    duration: 45
    fig_height: 6
    fig_width: 10
  beamer_presentation:
    fonttheme: professionalfonts
---

## Visualization

It's impossible to overstate the importance of visualization in data analysis.

* Helps us explore data
* Suggest a model
* Assess the validity of a model and its parameters
* Vital for a non-technical audience

## Visualization in R

4 plotting engines (at least)

* base plotting system
* lattice
* ggplot2
* rCharts

We'll look at the base plotting system now and ggplot2 after lunch.

## Common geometric objects

* scatter
* line
* hist
* density
* boxplot
* barplot
* dotplot

plot is the most basic graphics command. There are several dozen options that you can set. Spend a lot of time reading the documentation and experimenting.

##

Open your first script.

## A basic scatter plot
```{r }
source("../inst/scripts/BasicScript.R")
plot(X1, Y, pch=19)
```

## Add lines
The functions 'lines' and 'points' will add (wait for it) lines and points to a pre-existing plot.
```{r  }
plot(X1, Y, pch=19)
lines(X1, yHat)
```

## 
Can everyone generate a scatter plot?

## Histogram
```{r fig.height=4.5}
hist(e)
```

## Density plot
```{r fig.height=4.5}
plot(density(e))
```

## Boxplot
```{r fig.height=4.5}
boxplot(e, pch=19)
```

## Plotting a formula
```{r fig.height=4.5}
plot(Y ~ X1, pch=19)
```

## Emphasizing outliers
```{r fig.height=4.5}
colors = ifelse(abs(e) > 1.0, "red", "black")
plot(Y ~ X1, pch=19, col=colors)
```

## Other ways to emphasize outliers
```{r fig.height=4.5}
plot(Y ~ X1, pch=19)
lines(X1, yHat, lwd=2)
lines(X1, yHat+1, lty="dotted", lwd=0.5)
lines(X1, yHat-1, lty="dotted", lwd=0.5)
```

## Exercise

* Load the COTOR2 data from the raw package.
* Create a histogram, kernel density plot and box plot for the claims data

## Answer
```{r }
library(raw)
data(COTOR2)
hist(COTOR2)
boxplot(COTOR2)
plot(density(COTOR2))
plot(density(log(COTOR2)))
hist(log(COTOR2))
hist(COTOR2, breaks=80)
```

## Resources

* [Nathan Yau - FlowingData.com](http://flowingdata.com/)
* [Stephen Few - PerceptualEdge.com](http://www.perceptualedge.com/)
* [Edward Tufte - edwardtufte.com](http://www.edwardtufte.com/tufte/)
* [junkcharts.typepad.com](http://junkcharts.typepad.com/)